Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-06-28 paper.tex
.

The example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}). Some aggregated indicators will not take into account positions and trajectories of each agent (such as segregation in the Schelling model) but others, as in the case of spatial patterns of accessibility in a system of cities, fully capture the path-dependency and may therefore be highly dependent of the initial spatial configuration. It is not clear for example what shifted the economical and political capital of France from Lyon to Paris in the early Middle Age, some assumptions being the reconfiguration of trade patterns from South to North of Europe and thus an increased centrality for Paris due to its spatial position: the bifurcation induced by socio-economic and political factors took a deep significance with worldwide repercussions until today when magnified by the spatial configuration. \comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ? - je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.} \comment[MLT]{L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?}

@@

\[
d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
\]

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Model Exploration Workflows as an (meta-)sensitivity analysis method}

\comment[JR]{ca serait interessant d'inserer un paragraphe specifique sur Openmole, le multimodeling, les workflows modulables etc, comme outil très puissant pour ce qu'on veut faire : ca permet de rendre transparent les analyses de sensitivite mais egalement les meta-analyses de sensitivite (d'ou le titre).}

@@

More precisely, we explore the basic parameter space of the model, which is the population of agents $P$, the minimal ressource capacity

% summary stats
%    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
%0.009408 0.084490 1.266000 0.930400 1.574000 1.858000

@@

%relativedistance_metaparams.pdf
%relativedistance_morphspace.pdf

%phasediagram_id0_maxSugar110.png
%phasediagram_id27_maxSugar110.png

% phase diagrams
%          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
% id=27 : 0.7913103    2.376837   0.1440293 157.4147 4852.746
% id=0 : 2.562398    3.753032   0.1316788 128.4632 4753.983
% maxSugar = 110

>>>>

The example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}). Some aggregated indicators will not take into account positions and trajectories of each agent (such as segregation in the Schelling model) but others, as in the case of spatial patterns of accessibility in a system of cities, fully capture the path-dependency and may therefore be highly dependent of the initial spatial configuration. It is not clear for example what shifted the economical and political capital of France from Lyon to Paris in the early Middle Age, some assumptions being the reconfiguration of trade patterns from South to North of Europe and thus an increased centrality for Paris due to its spatial position: the bifurcation induced by socio-economic and political factors took a deep significance with worldwide repercussions until today when magnified by the spatial configuration. \comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ? - je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}[(MLT)L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?]

@@

\begin{equation}\label{eq:phase-distance}
d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Model Exploration Workflows as an meta-sensitivity analysis method}

The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows. The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in~\cite{openshaw1983data}, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities: \cite{cottineau2016back} claims for a renewed approach on multi-modeling. Coupling models as we do answers to similar questions. In that stream of research, the model exploration platform OpenMole~\cite{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.

@@

More precisely, we explore a grid of a basic parameter space of the model, which three dimensions are the population of agents $P\in \left[10;510\right]$, the minimal initial agent ressource $s_{-}\in \left[10;100\right]$ and the maximal initial agent ressource $s_{+}\in \left[110;200\right]$. Each parameter is binned into 10 values, giving 1000 parameter points. We run 50 repetitions for each configuration, what yield reasonable convergence properties. The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS. We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model. We mesure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup (see Fig.~\ref{fig:sugarscape-distance} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability. Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each. Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.

% summary stats
%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
% 0.08909 0.19790 1.52200 1.29600 2.16400 2.98100

We obtain a very strong sensitivity to initial conditions, as the distribution of the relative distance to reference across grids ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30. It means that in average, the model is more sensitive to meta-parameters than to parameters, and the relation variation can reach a factor of 3. We plot in Fig.~\ref{fig:sugarscape-distance} their distribution in a morphological space. The reduced morphological space is obtained by computing 4 raw indicators of urban form, namely Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and contextualization), and by reducing the dimension with a principal component analysis for which we keep the first two components (92\% of cumulated variance). The first measures a ``level of sprawl'' and of scattering, whereas the second measures aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation. It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance. In terms of model processes, it shows that congestion mechanisms induce rapidly higher levels of inequality.

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/relativedistance_metaparams}\\
\includegraphics[width=0.5\textwidth]{figures/relativedistance_morphspace}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} (Top) Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and diffusion ($\beta$, strength of diffusion process). (Bottom) Relative distance as a function of two first principal components of the morphological space (see text). Red point correspond to the reference spatial configuration. Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
\label{fig:sugarscape-distance}
\end{figure}
%%%%%%%%%%%%%

% phase diagrams -> ok well different qualitatively
%          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
% id=27 : 0.7913103    2.376837   0.1440293 157.4147 4852.746
% id=0 : 2.562398    3.753032   0.1316788 128.4632 4753.983
% maxSugar = 110

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.24\textwidth]{figures/phasediagram_id27_maxSugar110}
\includegraphics[width=0.24\textwidth]{figures/phasediagram_id0_maxSugar110}
\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$. (Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
\label{fig:sugarscape-phasediagrams}
\end{figure}
%%%%%%%%%%%%%

We now check the sensitivity in terms of qualitative behavior of phase diagrams. We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value. These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance}. The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories. The two examples have not only a very distant baseline inequality (the ceil of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values. The process is thus completely inverted, what would have significant impacts if one tried to schematize policies from this model. This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
